Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 Rmd/00-c-declaration.Rmd
typo fixes

This thesis includes 4 publications, two of which have been published and two 
which have not been submitted yet [@Lee2019; @Lee2020-ie].
As the core theme of my thesis is the development of software
interfaces for biological data analysis, and given the collaborative nature
of statistical computing and bioinformatics research; all of the included
papers in this thesis reflect and acknowledge the contributions of my 
co-authors. Of special note is chapter \@ref(ch:intron), which reflects my 
contribution (in the form of a software package) to the submitted manuscript entitled 
_Covering all your bases: incorporating intron signal from RNA-seq data_
[@Lee2019-mf], and includes new data analyses not found in the paper. 
The following table details the publications, including my and my fellow co-authors contributions:

>>>>

This thesis includes 4 publications, two of which have been published and two 
which have not been submitted yet [@Lee2019; @Lee2020-ie].
As the core theme of my thesis is the development of software
interfaces for biological data analysis, and given the collaborative nature
of statistical computing and bioinformatics research; all of the included
papers in this thesis reflect and acknowledge the contributions of my 
co-authors. Of special note is chapter \@ref(ch:intron), which reflects my 
contribution (in the form of a software package) to the published manuscript entitled 
_Covering all your bases: incorporating intron signal from RNA-seq data_
[@Lee2019-mf], and includes new data analyses not found in the paper. 
The following table details the publications, including my and my fellow co-authors contributions:

### 2020-10-28 Rmd/03-fluentGenomics.Rmd
correct path
typo fixes

```{r workflow, fig.cap = "(ref:workflow)"}
knitr::include_graphics(here::here("img", "workflow-fluentgenomics.png"))
```

@@

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se).
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the **tidyverse** design principles @tidyverse.

```{r se, fig.cap = "(ref:se)"}
knitr::include_graphics(here::here("img", "se-digram.png"))
```

(ref:se): A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples. The rows have their own accessor functions which contains additional data about the feature of interest, in this `rowRanges()` returns a _GRanges_  where rows are the genomic coordinates for each gene measured in the assay, while `colData()` returns a _DataFrame_  where rows contain information about the samples.

@@

(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis. The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline. The x-axis gives a sense of the total expression of the gene.

>>>>

```{r workflow, fig.cap = "(ref:workflow)"}
knitr::include_graphics(here::here("img", "workflow-fluentgenomics.pdf"))
```

@@

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se)).
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the **tidyverse** design principles @tidyverse.

```{r se, fig.cap = "(ref:se)"}
knitr::include_graphics(here::here("img", "se-diagram.pdf"))
```

(ref:se) A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples. The rows have their own accessor functions which contains additional data about the feature of interest, in this `rowRanges()` returns a _GRanges_  where rows are the genomic coordinates for each gene measured in the assay, while `colData()` returns a _DataFrame_  where rows contain information about the samples.

@@

(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes  that have an adjusted *p-value* below 0.01 are colored blue.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis. The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline. The x-axis gives a sense of the total expression of the gene.

### 2020-10-28 Rmd/05-embeddings.Rmd
typo fixes

There are some limitations
in using the **liminal** interface for larger datasets. First,
t-SNE avoids the crowding problem, points are separated into distinct regions
on the display canvas, while for the tour, points are concentrated in the
centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 
Second, as $n$ increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas. We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres. Alternative displays via statistical transformations could also mitigate the need to show all of the data. Recent work
by @Laa2020-wr is a promising area for further investigation, as well as
work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].

>>>>

There are some limitations
in using the **liminal** interface for larger datasets. First,
t-SNE avoids the crowding problem, points are separated into distinct regions
on the display canvas. For the tour, points are concentrated in the
centre of the projection and become difficult to see. We have recently proposed a simple non-linear transformation for the tour called a sage tour that aims to fix this problem [@Laa2020-vk]. 
Second, as $n$ increases both the embedding view and tour view become harder to read due to over-plotting, while the interactivity and animation become slower as there is more data passing from the server to the client.
For the tasks we have looked at in this paper, where shape and density are important to the analyst, we think that better displays and sub-sampling strategies are more useful than being able to look at every single point on the canvas. We showed in our single cell clustering case study that doing a weighted sample based on cluster membership still allowed us to get a sense of relative cluster orientation, however there are alternative sampling approaches that could be applied, like selecting points close to the cluster centres. Alternative displays via statistical transformations could also mitigate the need to show all of the data. Recent work
by @Laa2020-wr is a promising area for further investigation, as well as
work from topological statistics [@Rieck2017-kk; @Genovese2017-iq].
